Modelling the emergence of whisker barrels

### 2020-06-02 paper/briefpaper/brief.tex
Adds expanded section on how the gammas were set

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
  \color{blue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}
  \color{black}
\end{equation}
%
\cmnt{which contributes to the \emph{flux of axonal branching} as an
  additional diffusion-like term which makes the branching density for a given
  projection `flow away' from regions} where the branches of other projections
become dense. \cmnt{$\epsilon$ parameterizes the strength of this competitive
  interaction.} Note that this operation is local to individual afferent
projections.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. As such, two orthogonal linear thalamic
gradients were defined, from which 41 pairs of $\gamma$ values were sampled,
at the coordinates of 41 barreloid centers recreated from
\cite{haidarliu_size_2001} (Fig.\,\ref{fig:main}B).

A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a(\mb{x},0)$ and $c(\mb{x},0)$, a clear Voronoi-like tesselation emerged
(Fig.\,\ref{fig:main}C; see Movie S1). A reduction in the Honda delta metric (see
\citealp{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged
within \cmnt{$\approx 20000$} iterations (Fig.\,\ref{fig:main}D). A measure of the difference
in \cmnt{barrel area} between real and simulated barrels revealed a strong correspondence
(see Fig.\,\ref{fig:main}D, squares). \cmnt{In a chosen region, the ratio of \emph{the connection density
  of the TC type $i$ which makes the densest connections} to \emph{the sum of all
  connection densities} gives a measure of the localization of the connection
density. We computed this measure across the entire barrel domain and named it $\omega$
(see Fig.\,\ref{fig:main}D, grey hexagons). $\omega$ increases during the first $20000$ simulation
steps and converges on a stable value. Fig.\,\ref{fig:main}E shows
the localization measure plotted as a luminance map, giving a more graded
impression of the structure which develops in the system.}

@@

To further investigate the interplay of genetic and intrinsic mechanisms we
simulated a seminal barrel duplication experiment
\citep{shimogori_fibroblast_2005}, first by solving on a cortical domain
comprising two separate, identically shaped cortical boundaries. In this case
one distinct cortical field per thalamic projection emerged, within one of the
two boundaries only, i.e., no duplication. However, merging the two fields to
create an extended mirror-symmetric boundary shape, with an anterior-posterior
guidance field reflected at the join, gave rise to two mirror-symmetrical
barrel fields comprising 2N barrels, i.e., two cortical fields for each
thalamic projection (Fig.\,\ref{fig:fgf8}). Together these results suggest that by the
misexpression of Fgf8, \cite{shimogori_fibroblast_2005}
effectively created one large barrel field rather than two distinct cortical
fields.

>>>>

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching,
\cmnt{whilst ensuring that $a_i$ is conserved over time,} we
redefined
%
\begin{equation} \label{eq:comp}
  \color{blue}
  \chi_i(\mb{x}, t) = \frac{\epsilon a_i}{N-1} \nabla \textstyle\sum_{j\ne i}^{N}{a_j}
  \color{black}
\end{equation}
%
\cmnt{which contributes to the \emph{flux of axonal branching} as an
  additional diffusion-like term which makes the branching density for a given
  projection diffuse away from regions} where the branches of other projections
become dense. \cmnt{$\epsilon$ parameterizes the strength of this competitive
  interaction.} Note that this operation is local to individual afferent
projections.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. \cmnt{Thus, we suggest that there exist a
  pair of signalling molecules whose expression varies linearly (with
  orthogonal gradients) across the thalamus, analgous to the ephrins, whose
  existence is well known in the cortex. Thalamocortical axons growing from a
  region in the thalamus with high expression of one of these molecules will
  interact with a corresponding cortical signalling molecule more strongly
  than an axon emanating from a region with low expression of the thalamic
  signal.}  \cmnt{We defined two orthogonal, linear gradients oriented in the
  anterior-posterior and medial-lateral directions and used the coordinates of
  41 barreloid centers recreated from \cite{haidarliu_size_2001},
  corresponding to the macrovibrissae, to populate 41 pairs of $\gamma$
  values.}  (Fig.\,\ref{fig:main}B). \cmnt{In other words, we simply scaled
  the barreloid coordinates in an anterior-posterior/medial-lateral coordinate
  system to create the $\gamma$ values. Note that positive \emph{and} negative
  $\gamma$ values were generated; the origin of the coordinate system was
  located at the center of the thalamic barreloid field. This prevented an
  overall drift of the cortical pattern away from the centroid of the barrel
  field and may explain why coordinate-giving transcription factors and the
  ephrins are often found as pairs acting in opposing directions.}
\cmnt{We concentrated on macrovibrissae (excluding microvibrissae) because it
  is easy to identify their corresponding structures both in the thalamus and
  in the cortex. We also excluded the 5 rhinal whiskers because their cortical
  representation is in a separate island, distinct, and somewhat removed from
  the representation of the macrovibirissae.}

A cortical boundary enclosing 41 \cmnt{macrovibrissae} barrels was traced from
a cytochrome oxidase stain from \cite{zheng_signal_2001} \cmnt{(using the
  authors' original data)}, and Eqs.~\ref{eq:dc}--\ref{eq:comp} were solved
for $N=$ 41 projections on the resulting domain, using $M=2$ linear signalling
gradients aligned with the anterior-posterior and medial-lateral axes. From
random initial conditions for $a(\mb{x},0)$ and $c(\mb{x},0)$, a clear
Voronoi-like tesselation emerged (Fig.\,\ref{fig:main}C; see Movie S1). A
reduction in the Honda delta metric (see \citealp{senft_mouse_1991}) confirmed
that a `good' Voronoi pattern emerged within \cmnt{$\approx 20000$} iterations
(Fig.\,\ref{fig:main}D).  A measure of the difference in \cmnt{barrel area}
between real and simulated barrels revealed a strong correspondence (see
Fig.\,\ref{fig:main}D, squares). \cmnt{In a chosen region, the ratio of
  \emph{the connection density of the TC type $i$ which makes the densest
    connections} to \emph{the sum of all connection densities} gives a measure
  of the localization of the connection density. We computed this measure
  across the entire barrel domain and named it $\omega$ (see
  Fig.\,\ref{fig:main}D, grey hexagons). $\omega$ increases during the first
  $20000$ simulation steps and converges on a stable
  value. Fig.\,\ref{fig:main}E shows the localization measure plotted as a
  luminance map, giving a more graded impression of the structure which
  develops in the system.}

@@

To further investigate the interplay of genetic and intrinsic mechanisms we
simulated a seminal barrel duplication experiment
\citep{shimogori_fibroblast_2005}, first by solving on a cortical domain
comprising two separate, identically shaped cortical boundaries. In this case
one distinct cortical field per thalamic projection emerged, within one of the
two boundaries only, i.e., no duplication. However, merging the two fields to
create an extended mirror-symmetric boundary shape, with an anterior-posterior
guidance field reflected at the join, gave rise to two mirror-symmetrical
barrel fields comprising 2N barrels, i.e., two cortical fields for each
thalamic projection (Fig.\,\ref{fig:fgf8}). Together these results suggest that by the
misexpression of Fgf8, \cite{shimogori_fibroblast_2005}
effectively created one large barrel field rather than two distinct cortical
fields. \cmnt{Add something about the slightly off-vertical arrangement of the
  fields.}
